---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rconnection

<!-- badges: start -->
[![R-CMD-check](https://github.com/coolbutuseless/rconnection/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/rconnection/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`rconnection` demonstrates how to implement a custom connection in R.

This custom connection (`vfile()`) works similarly to the `file()` connection,
and all code for this implementation is included in this repository.

This code is licensed under the MIT License - feel free to share, fork and hack!

## Installation

You can install the development version of rconnection from [GitHub](https://github.com/coolbutuseless/rconnection) with:

``` r
# install.packages("devtools")
devtools::install_github("coolbutuseless/rconnection")
```

## Using `vfile()` to read/write binary data

```{r example}
library(rconnection)

tmp <- tempfile()
ref <- as.raw(1:255)
writeBin(ref, vfile(tmp, verbosity = 1))
tst <- readBin(vfile(tmp, verbosity = 1),  raw(), 1000)
tst
```


## Using `vfile()` to read/write text data


```{r}
ref <- as.character(mtcars)
writeLines(ref, vfile(tmp, verbosity = 1))
tst <- readLines(vfile(tmp, verbosity = 1))
tst
```

